---
title: "432 Assignment 4 Answer Sketch"
author: "432 Staff"
date: "Due 2018-02-23 at 1 PM. Version: `r Sys.Date()`"
output:
  pdf_document:
    toc: TRUE
    number_sections: TRUE
  html_document:
      code_folding: show
      toc: TRUE
      toc_float: TRUE
      number_sections: TRUE
---

## Setup

```{r setup, message = FALSE}
knitr::opts_chunk$set(comment=NA)

library(rms)
library(skimr)
library(broom)
library(tidyverse)

skim_with(numeric = list(hist = NULL), 
          integer = list(hist = NULL))

bird <- read.csv("bird.csv") %>% tbl_df
```

# Question 1 (60 points)

> You will fit a logistic regression model to address the key research question here, which is: "After age, sex, socio-economic status and smoking have been controlled for, is there an additional risk associated with keeping a bird as a pet?"

> You will need to:

> 1. specify an appropriate model for the data, then 
> 2. evaluate the quality of that model, appropriately,
> 3. and then provide an estimate (odds ratio with associated 95% confidence interval and careful interpretation) that addresses the research question above directly, then state your conclusion about whether this additional risk exists, and if so, how large is it?

> Some specific suggestions:

> - Use complete English sentences, punctuated by (well-edited) critical statistical output. Include the code used to produce that output in your HTML file.
> - Focus your presentation on the things that are *most important* for your reader to see.
> - Feel free to fit the simplest possible model that meets the requirements of the question. 
> - Your model will need to include all of the effects that are supposed to be accounted for, but you need not fit complex interaction or other non-linear terms on the right-hand side of the model unless you choose to do so.
> - You will have multiple decisions to make about how best to fit and analyze your model. Describe those choices well in your response. 

## Tidying data

Before doing any analysis, let's make sure the data is amenable to being analyzed: 

```{r}
skim(bird)
```

There is no missingness. None of the values seem out of line.

## Specifying an appropriate model

The simplest possible model that meets the requirements of the problem would include only the main effects of our four controlling variables (age, sex, socio-economic status and smoking) as adjustors for the key predictor (`petbird`)'s effect on lung cancer. It's your decision how best to incorporate the two variables related to smoking. 

Before specifying our model, we might look at the Spearman $\rho^2$ plot...

```{r rho-squaredplot}
plot(spearman2(lungc ~ petbird + sex + ses + 
                   cigs + smokeyr + age, data = bird))
```

Looking at the Spearman $\rho^2$ plot, it may be worth using `smokeyr` rather than `cigs`, so we'll try that first. 

## Building Model 1 (using glm)

```{r m1}
model.1 <- glm(lungc ~ petbird + sex + ses + age + smokeyr,
               data = bird, family = binomial(logit))
summary(model.1)
```

And now, here are the estimated odds ratios and confidence intervals.

```{r m1_oddsratios}
exp(coef(model.1))
exp(confint(model.1))
```

### Tidying the Model 1 Results

A better way to store these may be:

```{r}
m1_coeffs <- tidy(model.1, exponentiate = TRUE, conf.int = TRUE)

m1_coeffs
```

### Describing the Odds Ratio for `petbird`

The odds ratio for lung cancer associated with `petbird` after adjustment for `age`, `sex`, `ses` (socio-economic status) and `smokeyr` (years of smoking) is `r m1_coeffs %>% filter(term == "petbird") %>% select(estimate) %>% round(., 2)` with 95% CI for that odds ratio of (`r m1_coeffs %>% filter(term == "petbird") %>% select(conf.low) %>% round(., 2)`, `r m1_coeffs %>% filter(term == "petbird") %>% select(conf.high) %>% round(., 2)`).

This means that the odds of lung cancer for someone who keeps a pet bird were `r m1_coeffs %>% filter(term == "petbird") %>% select(estimate) %>% round(., 2)` times higher than the odds of lung cancer for someone who did not keep a bird, assuming that these two people had the same values of `sex`, `ses`, `age` and `smokeyr`.

### Fitting Model 1 with `lrm` to estimate C, R^2^

We could also have fit model.1 with the `lrm` function from the `rms` package.

```{r lrm_model}
d <- datadist(bird)
options(datadist="d")

mod1_lrm <- lrm(lungc ~ petbird + sex + ses + age + smokeyr, 
             data=bird, x = TRUE, y = TRUE)
mod1_lrm
```

### Evaluating Model 1

The C statistic and R^2^ values are almost good.

### Visualizing Model 1's Predictions

Note that, by specifying `plogis`, we are telling R to hold several variables at their median values and give us the *probabilities* of our outcome rather than the coefficients or odds ratios. 

```{r}
ggplot(Predict(mod1_lrm, fun = plogis), 
       ylab = "Est. Pr(lung cancer)")
```

## An Alternative Model (Model 2)

An alternative model using `cigs` rather than `smokeyr` would be even more aggressive, yielding the following: 

```{r}
model.2 <- glm(lungc ~ petbird + sex + ses + age + cigs,
               data = bird, family = "binomial")

m2_coeffs <- tidy(model.2, exponentiate = T, conf.int = T)

m2_coeffs
```

### Looking at the C statistic and R^2^ for Model 2 

```{r}
mod2_lrm <- lrm(lungc ~ petbird + sex + ses + age + cigs, 
                data = bird, x = TRUE, y = TRUE)
mod2_lrm
```

### Visualizing Model 2's Predictions

Here's the set of plots for Model 2. 

```{r}
ggplot(Predict(mod2_lrm, fun = plogis), 
       ylab = "Est. Pr(lung cancer)")
```

### Describing the Odds Ratio for `petbird` in Model 2

The odds ratio for lung cancer associated with `petbird` after adjustment for `age`, `sex`, `ses` (socio-economic status) and `cigs` (cigarettes smoked per day) is `r m2_coeffs %>% filter(term == "petbird") %>% select(estimate) %>% round(., 2)` with 95% CI for that odds ratio of (`r m2_coeffs %>% filter(term == "petbird") %>% select(conf.low) %>% round(., 2)`, `r m2_coeffs %>% filter(term == "petbird") %>% select(conf.high) %>% round(., 2)`).

This means that the odds of lung cancer for someone who keeps a pet bird were `r m2_coeffs %>% filter(term == "petbird") %>% select(estimate) %>% round(., 2)` times higher than the odds of lung cancer for someone who did not keep a bird, assuming that these two people had the same values of `sex`, `ses`, `age` and `cigs`.

### Evaluating Model 2

The C statistic and R^2^ statistic adjusting for `cigs` turn out to be a bit worse than the model with `smokeyr`.


## Yet another approach (Model 3)

An alternative model using `cigs` and`smokeyr` to calculate *pack years* `packyr` by multiplying # packs of cigarettes smoked per day (20 cigarettes = 1 pack) by the number of years the person has smoked to account for smoking history is worth a look, too.

```{r}
bird <- bird %>%
    mutate(packs = cigs/20,
           packyr = packs*smokeyr)
```

```{r}
model.3 <- glm(lungc ~ petbird + sex + ses + age + packyr,
               data = bird, family = "binomial")

m3_coeffs <- tidy(model.3, exponentiate = T, conf.int = T)

m3_coeffs
```

### Looking at the C statistic and R^2^ for Model 2 

```{r}
mod3_lrm <- lrm(lungc ~ petbird + sex + ses + age + packyr, 
                data = bird, x = TRUE, y = TRUE)
mod3_lrm
```

### Describing the Odds Ratio for `petbird` in Model 3

The odds ratio for lung cancer associated with `petbird` after adjustment for `age`, `sex`, `ses` (socio-economic status) and `packyr` (smoking pack-years) is `r m3_coeffs %>% filter(term == "petbird") %>% select(estimate) %>% round(., 2)` with 95% CI for that odds ratio of (`r m3_coeffs %>% filter(term == "petbird") %>% select(conf.low) %>% round(., 2)`, `r m3_coeffs %>% filter(term == "petbird") %>% select(conf.high) %>% round(., 2)`).

This means that the odds of lung cancer for someone who keeps a pet bird were `r m3_coeffs %>% filter(term == "petbird") %>% select(estimate) %>% round(., 2)` times higher than the odds of lung cancer for someone who did not keep a bird, assuming that these two people had the same values of `sex`, `ses`, `age` and `packyr`.

### Evaluating Model 3

The C statistic and R^2^ statistic adjusting for `packyr` turn out to be in between our previous two models.

### Visualizing Model 3's Predictions

Here's the set of plots for Model 3. 

```{r}
ggplot(Predict(mod2_lrm, fun = plogis), 
       ylab = "Est. Pr(lung cancer)")
```

# Question 2 (40 points)

> 1. First, in 2-4 complete English sentences, I want you to specify, using your own words and complete English sentences, the most useful and relevant piece of advice you took away from reading Jeff Leek's *How To Be A Modern Scientist*. Please provide a reference to the section of the book that provides this good advice. (For those of you who can more easily find things to gripe about in the book, don't worry - you will get that chance down the line.) 

> 2. Then, in an essay of 4-8 additional sentences, describe why this particular piece of advice was meaningful or useful for you, personally, and how it will affect the way you move forward. You are encouraged to provide a specific example of a past or current scientific experience of yours that would have been (or is being) helped by this new approach or idea. Why is this idea important and worth sharing?

**We don't write answer sketches for essay questions.**